Parametricity for Primitive Nested Types and GADTs

### 2020-12-16 free-theorems-nesteds/20-mscs.tex
Sections 5-8 for proofreading

Note that since the body F of a type $(\mu \phi. \lambda
\ol\alpha. F)\ol{G}$ can only be functorial in $\phi$ and the
variables in $\ol{\alpha}$, the representation of
$\mathit{List}\,\alpha$ as the ADT $\mu \beta. \,\onet + \alpha \times
\beta$ cannot be functorial in $\alpha$. By contrast, if
$\mathit{List}\,\alpha$ is represented as the nested type $(\mu
\phi. \lambda \beta.\,\onet + \beta \times \phi \beta)\,\alpha$ then
we can choose $\alpha$ to be a functorial variable or not when forming
the type. This observation holds for other ADTs as well; for example,
if $\mathit{Tree}\,\alpha\,\gamma = \mu \beta. \alpha + \beta \times
\gamma \times \beta$, then $\alpha, \gamma; \emptyset \vdash
\mathit{Tree}\,\alpha\,\gamma$ is well-formed, but $\emptyset; \alpha,
\gamma \vdash \mathit{Tree}\,\alpha\,\gamma$ is not. And it also
applies to some non-ADT types, such as $\mathit{GRose}\,\phi\,\alpha =
\mu \beta. \onet + \alpha \times \phi\beta$, in which $\phi$ and
$\alpha$ must both be non-functorial variables.  It is in fact
possible to allow ``extra'' $0$-ary functorial variables in the body
of $\mu$-types (functorial variables of higher arity are the real
problem), which would allow the first-order representations of ADTs to
be functorial. However, doing this requires some changes to the
formation rule for $\mu$-types, as well as the delicate threading of
some additional conditions throughout the model construction.  But
since we can always use an ADT's (semantically equivalent)
second-order representation when functoriality is needed, disallowing
such ``extra'' variables does not negatively impact the expressivity
of our calculus. We therefore pursue the simpler syntax here.
%The representation of $\mathit{List}\,\alpha$ as the ADT $\mu
%\beta. \onet + \alpha \times \beta$ cannot be functorial in
%$\alpha$. By contrast, its representation as the nested type $(\mu
%\phi. \lambda \beta.\,\onet + \beta \times \phi \beta)\,\alpha$ can be
%considered either functorial in $\alpha$ or not, according to whether
%$\alpha$ is in the functorial or non-functorial context in which this
%type is formed. If $\emptyset;\alpha \vdash \mathit{List} \,\alpha$,
%then $\vdash \Nat^\alpha \onet\, (\mathit{List}\,\alpha)$ is
%well-formed; if $\alpha;\emptyset \vdash \List \,\alpha$, then it is
%not.  Similarly, if $\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\,
%\alpha + \beta \times \gamma \times \beta$, then $\gamma;\emptyset
%\vdash \Nat^\alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \,
%\alpha\, \gamma)$ is well-formed, but $\emptyset;\gamma \vdash
%\Nat^\alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\,
%\gamma)$ is not. The body $F$ of a type $(\mu \phi.\lambda
%\ol\alpha.F)\ol G$ can only be functorial in $\phi$ and the variables
%in $\ol\alpha$. As a result, both $\phi$ and the variables in
%$\ol\alpha$ must be non-functorial in $\mathit{GRose}\, \phi \, \alpha
%= \mu\beta . \alpha \times \phi \beta$, $\mathit{GRose}\,\phi\,\alpha$
%cannot be the (co)domain of a $\Nat$-type binding any variables, and
%no type of the form $(\mu \phi. \lambda
%\alpha. \mathit{GRose}\,\phi\,\alpha)\tau$ is well-formed in our
%system.
%These restrictions ensure that the set and relational interpretations
%of types in Section~\ref{sec:type-interp} are functors. In
%Section~\ref{sec:conclusion} we discuss how more expressive types
%might be accommodated.

@@

Interestingly, even some recursive functions of a single proper nested
type --- e.g., a reverse function for bushes --- cannot be expressed
as folds because the algebra arguments needed to define them are again
recursive functions with types of the same problematic form as the
type of, e.g., a zip function for perfect trees.  Expressivity of
folds for nested types has long been a vexing issue, and this is
naturally inherited by our calculus. Adding more expressive recursion
combinators could help, but since this is orthogonal to the issue of
parametricity in the presence of primitive nested types we do not
consider it further here.

>>>>

Note that since the body F of a type $(\mu \phi. \lambda
\ol\alpha. F)\ol{G}$ can only be functorial in $\phi$ and the
variables in $\ol{\alpha}$, the representation of
$\mathit{List}\,\alpha$ as the ADT $\mu \beta. \,\onet + \alpha \times
\beta$ cannot be functorial in $\alpha$. By contrast, if
$\mathit{List}\,\alpha$ is represented as the nested type $(\mu
\phi. \lambda \beta.\,\onet + \beta \times \phi \beta)\,\alpha$ then
we can choose $\alpha$ to be a functorial variable or not when forming
the type. This observation holds for other ADTs as well; for example,
if $\mathit{Tree}\,\alpha\,\gamma = \mu \beta. \alpha + \beta \times
\gamma \times \beta$, then $\alpha, \gamma; \emptyset \vdash
\mathit{Tree}\,\alpha\,\gamma$ is well-formed, but $\emptyset; \alpha,
\gamma \vdash \mathit{Tree}\,\alpha\,\gamma$ is not. And it also
applies to some non-ADT types, such as $\mathit{GRose}\,\phi\,\alpha =
\mu \beta. \onet + \alpha \times \phi\beta$, in which $\phi$ and
$\alpha$ must both be non-functorial variables.  It is in fact
possible to allow ``extra'' $0$-ary functorial variables in the body
of $\mu$-types (functorial variables of higher arity are the real
problem), which would allow the first-order representations of ADTs to
be functorial. However, doing this requires some changes to the
formation rule for $\mu$-types, as well as the delicate threading of
some additional conditions throughout the model construction.  But
since we can always use an ADT's (semantically equivalent)
second-order representation when functoriality is needed, disallowing
such ``extra'' variables does not negatively impact the expressivity
of our calculus. We therefore pursue the simpler syntax here.

@@

Interestingly, even some recursive functions of a single proper nested
type --- e.g., a reverse function for bushes that is a true involution
--- cannot be expressed as folds because the algebra arguments needed
to define them are again recursive functions with types of the same
problematic form as the type of, e.g., a zip function for perfect
trees.  Expressivity of folds for nested types has long been a vexing
issue, and this is naturally inherited by our calculus. Adding more
expressive recursion combinators could help, but since this is
orthogonal to the issue of parametricity in the presence of primitive
nested types we do not consider it further here.
